### 2018-05-21 write-up/background.tex
Finish Background, continue Implementation

(* Owl (OCaml) *)
open Owl
let a = Mat.of_arrays [| [| 1.; 0. |]; [| 0.; 1. |] |]
let b = Mat.of_array [| [| 4.; 1. |]; [| 1.; 2. |] |]
let c = Mat.(a *@ b)
    \end{minted}
    \caption{Matrix Multiplication in Numpy (Python), Julia and Owl (OCaml).}\label{fig:mat_mul_copies}
\end{figure}

@@

\begin{figure}[tbp]
    \begin{minted}{fortran}
      program blasMatMul
      implicit none
      real*4 a(2,2), b(2,2), c(2,2)
C     External from BLAS
      external dgemm
C     Initialize in column major storage of Fortran
      data a/ 1,0,
     *        0,1/
      data b/ 4,1,
     *        1,2/
C                tfm  tfm  rowA colB K  alpha a lda  b  ldb beta c  ldc
      call dgemm('N', 'N', 2,   2,   2, 1.0,  a,  2, b, 2,  0.0, c, 2)
    \end{minted}
    \caption{One of \emph{several} BLAS (Fortran) routines for Matrix
    Multiplication.}\label{fig:fortran_blas}
\end{figure}

@@

\begin{figure}[tbp]
    \begin{minted}{c++}
#include <iostream>
#include <Eigen/Dense>
using namespace std;
int main()
{
    Eigen::Matrix2d a,b,c;
    a << 1, 0,
         0, 1;
    b << 4, 1,
         1, 2;
    c << 0, 0,
         0, 0;
    a * b; // new matrix
    c += a * b; // temporary for correctness in case of aliasing
    c.noalias() += a * b; // no temporaries
}
    \end{minted}
    \caption{Some examples of Matrix Multiplication in Eigen. Using expression
        templates (to be discussed later) and \emph{explicitly provided} aliasing
        information, Eigen can emit a single BLAS `dgemm'-like call for the last
        line, mirroring the Fortran example of
        Figure~\ref{fig:fortran_blas}.}\label{fig:cpp_eigen}
\end{figure}

@@

\section{TO DO: Proposed Solution}
A (E)DSL!
% https://github.com/AtheMathmo/rulinalg/blob/e0d8b76ddf8bbb5547cb9c98172e594605ab5fef/src/matrix/impl_ops.rs
% Which operand is overwritten determines scope. Would make sense to argue
% pros-cons of implicitness vs explicitness, emulating Rust's Affine with
% linear sugar.

% CODE EXAMPLES

@@

>>>>

(* Owl (OCaml) *)
open Owl
let a = Mat.of_arrays [|[|1.;0.|];[|0.;1.|]|]
let b = Mat.of_array [|[|4.;1.|];[|1.;2.|]|]
let c = Mat.(a *@ b)
    \end{minted}
    \caption{Matrix Multiplication in Numpy (Python), Julia and Owl (OCaml).}\label{fig:mat_mul_copies}
\end{figure}

@@

\begin{figure}[tbp]
    \begin{minted}[fontsize=\small]{fortran}
  program blasMatMul
  implicit none
  real*4 a(2,2), b(2,2), c(2,2)
C External from BLAS
  external dgemm
C Initialize in column major storage of Fortran
  data a/ 1,0,0,1/
  data b/ 4,1,1,2/
C            tfm  tfm  rowA colB K  alpha a lda  b  ldb beta c  ldc
  call dgemm('N', 'N', 2,   2,   2, 1.0,  a,  2, b, 2,  0.0, c, 2)
    \end{minted}
    \caption{One of \emph{several} BLAS (Fortran) routines for Matrix
    Multiplication.}\label{fig:fortran_blas}
\end{figure}

@@

It is important to note that should these annotations (in
Figures~\ref{fig:cpp_eigen} and~\ref{fig:cpp_ublas}) be wrong, the program's
behaviour is very likely to end up being undefined (like how \texttt{memcpy()}
for overlapping regions of memory is explicitly undefined in the POSIX and C
standards). Indeed, one of Fortran's strengths lies in assuming that references
cannot be aliased (with certain caveats) in more cases than C permits (this
informal, general statement comes with many nuances left for the interested
reader to pursue).

\begin{figure}[tp]
    \begin{minted}{c++}
#include <iostream>
#include <Eigen/Dense>
using namespace std;
int main()
{
    Eigen::Matrix2d a,b,c;
    a << 1, 0, 0, 1; b << 4, 1, 1, 2; c << 0, 0, 0, 0;
    a * b; // new matrix
    c += a * b; // temporary for correctness in case of aliasing
    c.noalias() += a * b; // no temporaries
}
    \end{minted}
    \caption{Some examples of Matrix Multiplication in Eigen. Using expression
        templates (to be discussed later) and \emph{explicitly provided} aliasing
        information, Eigen can emit a single BLAS `dgemm'-like call for the last
	line, mirroring the Fortran example of
	Figure~\ref{fig:fortran_blas}.}\label{fig:cpp_eigen}
\end{figure}

@@

\section{Proposed Solution}

My proposed solution to this dichotomy is a \emph{domain-specific language}
(DSL), called LT4LA (Linear Types for Linear Algebra). Although for expository
and testing purposes I have defined a concrete-syntax, a full implementation
would make use of a language's syntax-extension features (such as PPX for
OCaml) to \emph{embed} the DSL into the host language. Such an embedding is
straightforward but fairly tedious to implement. As a half-way point, I used
compile-time code generation to make the DSL's output available to OCaml for
testing and evaluation.

Let us have a look at a few examples of functions we can write with linear
types.  We can define the canonical factorial function
(Figure~\ref{fig:ltfla_factorial}) and sum over an array
(Figure~\ref{fig:ltfla_sumarray}).

\begin{figure}[tp]
    \inputminted[linenos, fontsize=\small]{ocaml}{../test/examples/factorial.lt}
    \caption{Factorial function in LT4LA.}\label{fig:ltfla_factorial}
\end{figure}

\begin{figure}[tp]
    \inputminted[linenos, fontsize=\small]{ocaml}{../test/examples/sum_array.lt}
    \caption{Summing over an array in LT4LA.}\label{fig:ltfla_sumarray}
\end{figure}

The syntax is intended to resemble OCaml's, apart from the spurious
\ltfla{`!'}s found here and there (they are annotations to show that we can
use a value more than once). In Figure~\ref{fig:ltfla_sumarray}, we see
\ltfla{row} has type (\ltfla{'x arr}). More detailed explanations of
\emph{what and why} will be given in Chapter~\ref{chap:impl}, but for now, it
is enough to know it means we can only \emph{read} from \ltfla{row}, and not
write to it. If we did try to write to or free \ltfla{row}, say by adding
another line as shown in Figure~\ref{fig:ltfla_univ}, would get a helpful error
message, as shown.

\begin{figure}[tp]
    \begin{minted}[linenos, firstnumber=7, fontsize=\small]{ocaml}
let row = row[i] := x1 in (* or *) let () = free row in
(* Could not show equality:                                        *)
(*     z arr                                                       *)
(* with                                                            *)
(*     'x arr                                                      *)
(*                                                                 *)
(* Var 'x is universally quantified                                *)
(* Are you trying to write to/free/unshare an array you don't own? *)
(* In test/examples/sum_array.lt, at line: 7 and column: 19        *)
    \end{minted}
    \caption{Attempting to write to or free a read only array in
    LT4LA.}\label{fig:ltfla_univ}
\end{figure}

Now suppose we were trying to square a matrix, using a `dgemm' like BLAS
routine which takes two read-only matrices and a third matrix it can write to
and performs $C := AB + C$. How would we use such a routine to square a matrix?
Surely this would break linearity, since $A$ would be have to be given to the
function twice, and we can only do so once?

We can use a special primitive called \ltfla{share} to produce read-only
\emph{aliases} of any matrix (including ones we cannot write to). We then pass
this into the function, and it works as expected. Once the squared matrix has
been obtained, we may not want the original any more, and thus decide to free
it. Before we do so, we must first \ltfla{unshare} any read-only aliases that
exist -- in this context, \ltfla{unshare} swallows them (the only primitive
which can return fewer read-only values than it receives) and gives back a
single, read-write handle.

\begin{figure}[tp]
    \begin{minted}[linenos, fontsize=\small]{ocaml}
let (a1, a2) = share _ a in
let ((a1, a2), c) = simple_dgemm _ a1 _ a2 c in
let a = unshare _ a1 a2 in
let () = free a in c
    \end{minted}
    \caption{Squaring a matrix in LT4LA.}\label{fig:ltfla_square}
\end{figure}

If we tried to free one of the read-only aliases instead of
\ltfla{unshare}-ing, then we would get the error show in
Figure~\ref{fig:ltfla_free}. Briefly, a \ltfla{z arr} is a read-write array,
everything else (\ltfla{'x arr} or \ltfla{_ s arr}) is read-only. The types of
\ltfla{free : z arr --o unit}, \ltfla{share : 'x arr --o 'x s arr * 'x s arr}
and \ltfla{unshare : 'x . 'x s arr --o 'x s arr --o 'x arr} are set up so that
you can only free something when you have read-write access to it, guaranteed
by linearity to be the only name in scope with this capability (aliases can
only be read-only).

\begin{figure}[tp]
    \begin{minted}[linenos, fontsize=\small]{ocaml}
let (a1, a2) = share _ a in
let ((a1, a2), c) = simple_dgemm _ a1 _ a2 c in
let () = free a1 in c
(* Error:                                                        *)
(* Could not show equality:                                      *)
(*     z arr                                                     *)
(* with                                                          *)
(*     z s arr                                                   *)
(*                                                               *)
(* Could not show z and z s are equal.                           *)
(* Are you trying to write to/free an array before unsharing it? *)
(* In test.lt, at line: 3 and column: 17                         *)
    \end{minted}
    \caption{Attempting to free a read-only alias of matrix.}\label{fig:ltfla_free}
\end{figure}

@@

\section{Summary}

I have given an \emph{intuitive} exposition of linear types with
\emph{fractional-capabilities}, emphasising small, but illustrative and
practical code examples, leaving \emph{what} is going on and \emph{why} it
works as details for the next chapter. I have shown that it is possible to
control aliasing and reading/writing capabilities \emph{statically} with a
\emph{type system} -- a programmer can be prevented from writing over and
freeing arrays they do not own whilst being able to safely create read-only
aliases when it makes sense to do so. Unlike Rust's more full-featured
borrow-checker, the next chapter will show that these features can be expressed
and implemented in a concise, \emph{high-level} and language-independent manner
to be packaged up, provided and used \emph{as a library}.

### 2018-05-21 write-up/implementation.tex
Finish Background, continue Implementation

\chapter{Implementation}

@@

    \item \textbf{Executable Artifacts}. A transpiler and a REPL are the main artifacts
        produced for this thesis.

    \item \textbf{Tests}. As mentioned before, almost all non-trivial functions
        have tests to check their behaviour. The output of the transpiler
        was also tested by having the build system generate code at OCaml code
        which could then be compiled and tested as usual.

@@

A full description of the core language can be found in
Appendix~\ref{chap:ott_spec}. To make a linearly-typed language usable, we need
some way of using values zero or more than once, as we would an intuitionistic
value. For this, we have in the type-expressions the !-constructor and in the
term-expressions the \textbf{many}-constructor. For soundness, I implemented
\emph{value-restriction} for this type system, only \emph{values} (anything not
an array, matrix or an expression that can be reduced) that use \emph{no linear
values} can be marked with !. This means we can freely have integers, booleans
and (appropriate) functions usable as normal. Recursive functions can only use
linear values that are passed in as arguments to the function.

% TODO Once I've implemented the <- syntax, this should just include
% ../test/examples/weighted_avg_infer.lt directly
\begin{figure}[tb]
    \begin{minted}[linenos]{ocaml}
let rec f (!i : !int)
        (!n : !int)
        (!x0 : !elt)
        (write : z arr)
        ('x)
        (weights : 'x arr)
        : z arr * 'x arr =
  if n = i then (write, weights) else
  let !w0 <- weights[0] in
  let !w1 <- weights[1] in
  let !w2 <- weights[2] in
  let !x1 <- write[i] in
  let !x2 <- write[i + 1] in
  let !x : !elt = w0 *. x0 +. w1 *. x1 +. w2 *. x2 in
  f (i + 1) n x1 (write[i] := x) _ weights
    \end{minted}

    \caption{A recursive function to perform a \emph{in-place} (simplified)
        1D-convolution (sliding weighted-average) using a read-only array of
        weights and a writeable array, originally containing the input
        array.}\label{fig:oned_conv}

@@

So, how does a typical programmer write code that uses a matrix or an array
more than once? Figure~\ref{fig:oned_conv} shows an example program that
executes a (simplified) 1D-convolution (sliding weighted-average) using a
read-only array of three weights. On line 1, \ltfla{let rec} introduces a
recursive function `f', followed by explicitly annotated formal parameters.
Intuitively, \ltfla{!id} allows \ltfla{id} to be used intuitionistically. The
types \ltfla{!int} and \ltfla{!elt} correspond to integers and element types
(for example, 64-bit floating-point).

Array and matrix types are parameterised by \emph{fractional-capabilities}.  A
fraction of 1 ($2^0$) represents complete ownership of a value; in
particular, this allows a programmer to write or free it. Creating an array
gives you ownership of it; the function \ltfla{array : !int --o z arr}
(where \ltfla{z} represents `0'). Once you have ownership of an
array/matrix, you can free it: \ltfla{free : z arr --o unit}.
Importantly, because a linear-value may only be used once, the array just freed
is \emph{out of scope} for following expressions, preventing use-after-free.
Ownership also enables you to write to the array: \ltfla{set : z arr
--o !int --o !elt --o z arr} (the syntax \ltfla{w[i] := j} is just
sugar for \ltfla{set w i j}). Here, linearity prevents accessing values
which represented the array \emph{before} the mutation.

Any fraction less than 1 (for simplicity, limited to $2^{-k}$ in this system,
for a positive integer $k$) represents read-only access. So, the \ltfla{'x}
represents a natural number (either a zero \ltfla{z}, variable \ltfla{'x} or a
successor (+1) of a natural number). Hence, you can read from (index) any array
\ltfla{get : 'x .  'x arr --o !int --o !elt * 'x arr} (the syntax \ltfla{let !v
<- w[i]} is just sugar for \ltfla{let (!v, w) = get _ w i}). Transparently
rebinding \ltfla{w} with the returned value means a program can appear to use
\ltfla{w} multiple times; the underscore is how a programmer tells the compiler
to automatically \emph{infer} the correct fractional-capability based on the
other arguments passed to the function. Fractions also permit aliasing:
\ltfla{share : 'x . 'x arr --o 'x s arr * 'x s arr}. However, given the types
of the primitives provided, these aliases can only be read from and not written
to. If you want to write to this array, you must \ltfla{unshare: 'x . 'x s arr
--o 'x s arr --o 'x arr} it until you are left with a value of type \ltfla{z
arr}, guaranteeing no other aliases exists.

Given this set-up, we now \emph{statically} have \emph{perfect} information
about aliasing and ownership of values in the program. We can only write to an
array when we own it; ownership guarantees no other aliases exist in scope at
the point of usage. In the next section, I show how this perfect information
can be used to write more natural-looking code using value-semantic expressions
which behave in precisely the way we intend it to. Now the programmer need not
resort to manually tracking and inserting \texttt{noalias} annotations; instead
they can let the loyal and tireless compiler do the heavy lifting.

@@

We have now arrived at the titular \emph{applications} of linear types. I will
show how, in addition to preventing the errors explained in the previous
section, we can take linear types one step further and apply it to the domain
of efficient compilation of matrix-expressions.

\begin{figure}[tb]
    \inputminted[linenos]{fortran}{kalman.f90}
    \caption{Kalman filter in Fortran 90 (Credit: Matthew
        Rocklin~\cite{rocklin_gist}).}\label{fig:fortran_kalman}
\end{figure}

In Figure~\ref{fig:fortran_kalman}, we see the difficulty of efficiently
implementing a \emph{Kalman filter}, a powerful and set of equations applicable
to a wide variety of problems. From the comments, we see that every variable is
annotated with the step/matrix expression that it will hold at some point
during the computation (an equivalent alternative, say in C++, could be to have
a meaningful name for each step/matrix expression and manually annotate/keep
track of which names alias the same location).

In contrast, Figure~\ref{fig:ltfla_kalman}, offers the advantages of
\begin{itemize}
    \item aliasing: labelling each step with a different, more meaningful variable name,
    \item easily spotting which resources are being passed in and which are
        allocated for the function (new/copy),
    \item unambiguously seeing \emph{when} and what values are freed;
\end{itemize}
and have the compiler automatically ensure the safety of each of the above by respectively
\begin{itemize}
    \item making it impossible to refer to steps/values which are no longer usable,
    \item ensuring all values are declared and \emph{initialised} correctly before they are used,
    \item checking no values are used after they are freed \emph{and} ensuring no values are leaked.
\end{itemize}

Indeed, an inexperienced programmer could take the  na\"ive approach of just
copying sub-expressions by default and then letting the compiler tell it which
copies are never used and removing them systematically until it type checks.
While it is not quite a black-box, push-button compilation of an expression, I
would argue that, it is just as easy (if not easier) to become familiar with as
Rust and its borrow-checker.

\begin{figure}[tb]

    \begin{minted}[linenos]{ocaml}
let !kalman
        ('s) (sigma : 's mat) (* n, n *)
        ('h) (h : 'h mat)     (* k, n *)
        ('m) (mu : 'm arr)    (* n    *)
        (r_1 : z mat)         (* k, k *)
        (data_1 : z arr)      (* k    *)
        : ('s mat * ('h mat * 'm arr)) * (z arr * z mat) =
    let ((!k, !n), h) = dimM _ h in
    let zeros = zeroM n n in
    let sigma_h = new [| sym (up, sigma) * h |] in
    let r_2 = [| h * sigma_h + r_1 |] in
    let mu_1 <- copy mu in
    let sigma_1 <- copyM sigma in
    let data_2 = [| h * mu + -1. * data_1 |] in
    let h_1 <- copyM h in
    let h_sol = [| sym (up, r) ^ -1 * h_1 ] in
    let data_sol = [| sym (up, r) ^ -1 * data_2 |] in
    let h_h1 = new [| h * h_sol |] in
    let h_data = new [| h * data |] in
    let h_h1_sigma = new [| sym (up, h_h1) * sigma |] in
    let mu_2 = [| sym (up, sigma) * h_data * mu_1 |] in
    (* non-linear use of sigma_c..? *)
    let sigma_c = [|  -1. * sym (up, sigma_c) * h_h1_sigma + sigma_c |] in
    (* frees... *)
    (mu_c , sigma_c)
    \end{minted}

    \caption{Kalman filter in LT4LA.}\label{fig:ltfla_kalman}

@@

So-called `matrix \emph{expressions}' are still `side-effecting'/consuming
linear values, and do not produce \emph{values} but a \emph{sequence of
(re-)bindings} which in turn dictate what values are still available/in-scope
after the computation.  As such, compilation cannot be done purely
compositionally via structural-induction on the expression language; it is most
concisely expressed via CPS, where the result of elaborating a matrix
expression is a function that takes as its argument the rest of the computation
expecting to use the result it has just made available under that requested
name.

\subsection{TO DO: Inference}

@@

\section{TO DO: Theory}

Fractional permissions, not proven soundness etc\. but compilation preserves
types as checked by OCaml.

>>>>

\chapter{Implementation}\label{chap:impl}

@@

    \item \textbf{Executable Artifacts}. A transpiler and a REPL are the main
        artifacts produced for this thesis. For evaluation, I implemented
        Kalman filters in Owl, LT4LA and CBLAS/LAPACKE and a benchmarking
        program to measure execution times.

    \item \textbf{Tests}. As mentioned before, almost all non-trivial functions
        have tests to check their behaviour. The output of the transpiler was
        also tested by having the build system generate OCaml code at compile
        time, which in turn could then be compiled and tested like handwritten
        OCaml code.

@@

A full description of the core language can be found in
Appendix~\ref{chap:ott_spec}. Its main features are intuitionistic values,
value-restriction, fractional-capabilities (inferrable at call sites),
if-expressions and recursion.

\subsection{Intuitionistic Values}

% Why, runtime representation, trade-offs etc.

To make a linearly-typed language usable, we need some way of using values zero
or more than once, as we would an intuitionistic value. For this, we have in
the type-expressions the !-constructor and in the term-expressions the
\ltfla{Many}-constructor and the \ltfla{let Many <id> = .. in .. } eliminator.
The idea behind the !-type is that the value uses no `resources'
(linearly-typed expressions). To start off with, it is enough to say that
anything which can be passed around by copying, will have a !-type. This
includes integers, elements and booleans. So \ltfla{3 : !int} and \ltfla{3. +.
4. : !elt}. However, all bindings are still linear by default, so to emulate
intuitionism, I desugar \ltfla{let !x = <exp> in <body>} to  \ltfla{let Many x
= <exp> in let Many x = Many (Many x) in <body>} (similarly for function
argument bindings). The reader can check using the rules in the Appendix that
this has the effect of moving \ltfla{x : !t} from the linear to the
intuitionistic environments, only if \ltfla{<exp> : !t}.

However, just that desugaring alone is not enough to prevent a user from taking
an array or matrix and moving it into the intuitionistic environments. Why?
There are certain situations in which we \emph{should not} use the \ltfla{Many}
constructor.  Consider the following code: \ltfla{let Many x = Many (array 5)
in <body>}; the expression \ltfla{array 5} uses no linearly-typed variables
from the linear environment. Although we could just reject types of the form
\ltfla{!(_ arr)} to fix this simple example, what about pairs \ltfla{let Many
xy = Many (3, array 5) in <body>}?  Ad-hoc pattern matching on the type cannot
account for all possible situations. With the last case, we can use \ltfla{xy}
as many times as we would like, destruct the pair to get the second component
and thus create \emph{distinct} read-write aliases to the same array.  Alas,
now arrays can be used intuitionistically and all the benefits of linearity are
lost.  Or are they?

\subsection{Value-Restriction}

Not quite, but to understand how we can fix this problem, we need to question
an assumption left implicit up until this point: what does \ltfla{Many} even
mean? What does it do at runtime? One option is to go down the C++ route and
make \ltfla{Many} act like a \ltfla{shared_ptr} and act as a runtime
reference-count for arrays. I chose to not go for this option because it went
against the \emph{explicitness} and \emph{predictability} that C and Fortran
have. It would make analysing when and what is allocated and freed more like
the higher-level languages I was trying to move away from.

My aim is to show linear types can be simple to understand and apply to linear
algebra, enough so that it can be grafted (in a limited way) on to existing
languages as a \emph{library}. In that spirit, the simplest thing that
\ltfla{Many} can mean at runtime is \emph{nothing}. This language construct is
translated into a standard OCaml language constructor of the form \ltfla{type
'a bang = Many of 'a [@@unboxed]}. The \emph{unboxed} annotation means
that the type and its constructor only exist for the purpose of type-checking
in OCaml; \emph{the runtime representation of values of type \ltfla{'a} and
\ltfla{'a bang} is exactly the same}.

With this understanding, our problem is that arrays and matrices are unlike
other values such as integers and elements because (under the OCaml hood)
calling a function with an array argument copies a \emph{pointer} to the array
rather than the array itself, instead of the \emph{value} itself. So, we can
start making a distinction, \emph{defining} elements, integers, booleans,
intuitionistic variables, units and lambda-expressions that capture no linear
variables as \emph{values} (since they cannot break referential-transparency)
and anything else (arrays, matrices, expressions which can be reduced, such as
function application and if-expressions) as not being `values'. If this sounds
familiar, it is because this is the same \emph{value-restriction} `trick' from
the world of polymorphic types applied to linearity instead. We then have the
rule that we can only use \ltfla{Many} on expressions that are defined to be
\emph{values} and \emph{use no linear variables}.

\subsection{Fractional-Capabilities and Inference}

Having understood and correctly implemented intuitionism, we now tackle
the problem of how to implement safe aliasing, that is taking a read-write
alias and splitting into read-only aliases, as well as the inverse.

Array and matrix types are parameterised by \emph{fractional-capabilities}.  A
fraction of 1 ($2^0$) represents complete ownership of a value; in particular,
this allows a programmer to write or free it. Creating an array gives you
ownership of it; the function \ltfla{array : !int --o z arr} (where \ltfla{z}
represents `0'). Once you have ownership of an array, you can free it:
\ltfla{free : z arr --o unit}.  Importantly, because a linear-value may only be
used once, the array just freed is \emph{out of scope} for following
expressions, preventing use-after-free.  Ownership also enables you to write to
the array: \ltfla{set : z arr --o !int --o !elt --o z arr} (the syntax
\ltfla{w[i] := j} is just sugar for \ltfla{set w i j}). Here, linearity
prevents accessing aliases which represented the array \emph{before} the
mutation.

Any fraction less than 1 (for simplicity, limited to $2^{-k}$ in this system,
for a positive integer $k$) represents read-only access. So, the \ltfla{'x}
represents a natural number (either a zero \ltfla{z}, variable \ltfla{'x} or a
successor (+1) of a natural number). Hence, you can read from (index) any array
\ltfla{get : 'x .  'x arr --o !int --o 'x arr * !elt} (the syntax \ltfla{let !v
<- w[i]} is just sugar for \ltfla{let (w, !v) = get _ w i}). Transparently
rebinding \ltfla{w} with the returned value means a program can appear to use
\ltfla{w} multiple times. The underscore is how a programmer tells the compiler
to automatically \emph{infer} the correct fractional-capability based on the
other arguments passed to the function. In conjunction with the requirement
that functions declarations need type-annotations for their arguments, this
allows a fractional-capability to be correctly inferred in any program.

Fractions exist solely to enable safe aliasing, via the primitive \ltfla{share
: 'x . 'x arr --o 'x s arr * 'x s arr}.  The two arrays returned (which happen
to just be the given array) can now only be read from and not written to.  If
you want to write to this array, you must \ltfla{unshare: 'x .  'x s arr --o
'x s arr --o 'x arr} it with other read-only aliases until you are left with a
value of type \ltfla{z arr}, guaranteeing no other aliases exists.

Given this set-up, we now \emph{statically} have \emph{perfect} information
about aliasing and ownership of values in the program. We can only write to an
array when we own it; ownership guarantees no other aliases exist in scope at
the point of usage. In Figure~\ref{fig:ltfla_kalman}, I show how this perfect
information can be used to write more natural-looking code using value-semantic
expressions which behave in precisely the way we intend it to.  Now the
programmer need not resort to manually tracking and inserting \texttt{noalias}
annotations; instead they can let the loyal and tireless compiler do the heavy
lifting.

\subsection{If-Expressions}

Because we do not know which way a condition will evaluate at run time, we must
guarantee the both branches use the same set of linear variables. Writing the
type-checker in a pure-functional monadic style paid off here because I could
now sandwich monadic values with state-adjustments either side of it. Given two
monadic values that represented type-checking two branches of an if-expression,
I could use the code in Figure~\ref{fig:same_resources} to easily save, reset
and compare the state either side of running those monadic values.

\begin{figure}[tp]
    \begin{minted}{ocaml}
let same_resources (wf_a, loc_a) (wf_b, loc_b) =
  let open Let_syntax in
  (* Save state *)
  let%bind {used_vars=prev; env=old_env; _} as state = get in
  (* Reset, run a, save state *)
  let%bind () = put { state with used_vars = empty_used } in
  let%bind res_a = wf_a in
  let%bind {used_vars=used_a; _} as state = get in
  (* Reset, run b, save state *)
  let%bind () = put { state with used_vars = empty_used; env = old_env } in
  let%bind res_b = wf_b in
  let%bind {used_vars=used_b; _} as state = get in
  (* Check if same resources *)
  let keys_a, keys_b = (* convert to (used_a, used_b) to sets *) in
  if Set.equal keys_a keys_b then
    (* merge used_vars and used_b environments *)
  else
    (* report differences *)
    \end{minted}

    \caption{Implementation of same\_resources helper method for type-checking
        if-expressions. Note the monadic style helped compose computations that
        affected the type-checker's state in a simple manner.}\label{fig:same_resources}

@@

\subsection{Functions and Recursion}

A non-recursive function may be used more than once if it does not refer to any
linear variables from the surrounding scope. So, we can desugar something like
\ltfla{let x = 3 in let !f (y : !int) = x + y in <body>} to \ltfla{let x = 3 in
let Many f =  Many (fun y : !int -> x + y) in  <body>}. Recursion is
slightly more complicated: we can desugar the factorial function
(Figure~\ref{fig:ltfla_factorial}) to \ltfla{let Many f = fix (f, x : !int,
if (*..*) : !int) in <body> }. However, \ltfla{fix}, like \ltfla{Many},
must also not use any linear variables from its surrounding scope.

@@

We have now arrived at the titular \emph{applications} of linear types. I will
show how, in addition to preventing the errors explained hitherto, we can take
linear types one step further and apply it to the domain of efficient
compilation of matrix-expressions.

\begin{sidewaysfigure}
    \inputminted[linenos, fontsize=\footnotesize]{fortran}{kalman.f90}
    \caption{Kalman filter in Fortran 90.}\label{fig:fortran_kalman}
\end{sidewaysfigure}

In Figure~\ref{fig:fortran_kalman}, we see the difficulty of efficiently
implementing a \emph{Kalman filter}, a powerful set of equations applicable
to a wide variety of problems. From the comments, we see that every variable is
annotated with the step/matrix expression that it will hold at some point
during the computation (an equivalent alternative, say in C++, could be to have
a meaningful name for each step/matrix expression and manually annotate/keep
track of which names alias the same location).

In contrast, Figure~\ref{fig:ltfla_kalman}, offers the advantages of
\begin{itemize}
    \item aliasing: labelling each step with a different, more meaningful variable name,
    \item easily spotting which resources are being passed in and which are
        allocated for the function (new/copy),
    \item unambiguously seeing \emph{when} and what values are freed;
\end{itemize}
and have the compiler automatically ensure the safety of each of the above by respectively
\begin{itemize}
    \item making it impossible to refer to steps/values which are no longer usable,
    \item ensuring all values are declared and \emph{initialised} correctly before they are used,
    \item checking no values are used after they are freed \emph{or} leaked.
\end{itemize}

Indeed, an inexperienced programmer could take the  na\"ive approach of just
copying sub-expressions by default and then letting the compiler tell it which
copies are never used and removing them systematically until it type-checks.
While it is not quite a black-box, push-button compilation of an expression, I
would argue that, it is just as easy (if not easier) to become familiar with as
Rust and its borrow-checker.

\begin{figure}[tb]
    \inputminted[linenos, fontsize=\small]{ocaml}{../test/examples/kalman.lt}
    \caption{Kalman filter in LT4LA.}\label{fig:ltfla_kalman}
\end{figure}

@@

So-called `matrix \emph{expressions}' are still `side-effecting'/consuming
linear variables, and do not produce \emph{values} but a \emph{sequence of
(re-)bindings} which in turn dictate what values are still available/in-scope
after the computation.  As such, compilation cannot be done purely
compositionally via structural-induction on the expression language; it is most
concisely expressed via CPS, where the result of elaborating a matrix
expression is a function that takes as its argument the rest of the computation
expecting to use the result it has just made available under that requested
name.

@@
